Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-12-08 MANUSCRIPT.tex
Writing
Methods - ML + Feature Selection

Results - Ungauged model validation + importance

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

This study develops a technique for estimating BFI in ungauged basins in
Arizona and determines the state's long-term mean BFI from 1991 to 2020.
Due to interannual variability in BFI, a 30-year period of record is
used to evaluate the characteristic base-flow dynamics of Arizona. Using
a machine learning model trained with the hydrogeologic characteristics
of surface water basins, we first determine annual BFIs for each year in
the 1991-2020 study period for ungauged river reaches and physiographic
regions of Arizona. Long-term mean BFI data for the study period are
then estimated. Regional patterns and trends in base flow of
instrumented sites are subsequently investigated. These results provide
novel analyses of low flow processes in Arizona\ldots\#\#\#\#\#\#

@@

The state of Arizona, located in the southwest United States, covers a
total area of 295,253 km\(^2\). Arizona is divided into two primary
physiographic provinces: the Colorado Plateau in the northeast, and the
Basin-and-Range region in the northwest and south. The Central Highlands
is a transition zone consisting of scattered basins separated by the
mountainous foothills of the Mogollon Rim. The Colorado Plateau is
dominated by high-elevation desert with an average elevation of 1,936
masl (6,352 ft). Average temperatures range from -6\(^\circ\)C
(20\(^\circ\)F) to 26\(^\circ\)C (80\(^\circ\)F) and it averages 580 mm
(23 in) of precipitation. The Basin-and-Range region has a semi-arid to
arid climate with an average elevation of 490 masl (1600 ft). The
average precipitation is 200 mm (8 in) annually with average
temperatures ranging from 15\(^\circ\)C (60\(^\circ\)F) to 43\(^\circ\)C
(110\(^\circ\)F).

@@

\includegraphics{images/StudyArea_20241126.png}

@@

Arizona's hydrology varies seasonally and spatially between its
physiographic regions. In the summer, localized and intense convective
precipitation events are driven by the North American Monsoon, while in
the winter, orographic precipitation comes from Pacific frontal systems
(Eastoe et al., 2019). While monsoonal precipitation can account for up
to 50\% of annual precipitation, evaporation and dry preceding soil
properties leads to most precipitation becoming runoff (Sheppard et al.,
2002). As such, 94\% of streams in Arizona are ephemeral or intermittent
(Levick et al., 2008). Much of the hydrology of Arizona is snow-melt
derived, driven by spring melt from the high-elevation Colorado Plateau
winter snowpack. While winter precipitation provides only 30\% of annual
averages, it provides the bulk of water for natural reservoirs (Sheppard
et al., 2002).

@@

Daily observed streamflow data obtained from the United States
Geological Survey (USGS) National Water Information System (NWIS) were
used in this study. Streamgages were selected depending on criteria to
ensure the applicability of each site. O'Donnell et al. (2016) found
that 8-10 years of calibration data were necessary to account for
climate variability in paired watershed studies in the region. Following
this criteria, a minimum of 10 years of record was chosen. Any years
missing more than 30 days of streamflow were removed from the possible
record period. This analysis focuses on natural, streamflow-influencing
recharge, excluding streamgages affected by regulation or diversions.
These gages were identified through annual reports on water data
published by the USGS (USGS, 2010). Finally, streamgages on the Colorado
River were excluded from the dataset on the basis that they represent
managed flows shared by Colorado River Compact states. Following these
parameters, 206 USGS streamgages with acceptable periods of record were
used in this study (Figure~\ref{fig-study-area}).

Watersheds across the United States are delineated by the USGS using a
hydrologically-defined network. This system delineates the country using
hierarchical hydrologic unit codes (HUCs), where each subsequent basin
includes the digits of the enclosing basin. Here, 8-digit HUCs (HUC 8s)
are used to divide Arizona into 84 sub-basins that are fully or
partially in the state (Figure~\ref{fig-study-area}). These HUC 8
sub-basins are analogous to medium-sized river basins and are defined by
surface water characteristics. \#\#\#\#\#\#

The annual precipitation and temperature data came from the PRISM
climate group at Oregon State University at a resolution of 4 km
(\url{https://prism.oregonstate.edu;} (Daly et al., 2008)). The PRISM
dataset provides valuable insights into regional climate in ungauged
regions and has been shown to perform well across the southwestern US
(Buban et al., 2020)\textbf{.} Instead of the water year, PRISM data
uses a calendar-year format, which was adopted for consistency in the
water balance. Although this may introduce challenges in the annual
estimates due to inter-annual snow storage, the use of long-term annual
averages is likely to reduce any potential errors (Reitz et al., 2017).

@@

Base flow was calculated using a single-parameter, recursive digital
filter technique from (Nathan et al., 1990). This base-flow separation
technique is based on a recursive digital filter used in signal analysis
that separates high-frequency signals (quickflow) from low-frequency
signals (base flow) (Lyne et al., 1979). This technique has been used in
(Arnold et al., 2000; Bloomfield et al., 2009; Santhi et al., 2008;
Singh et al., 2018). It takes the form of

@@

where \(b\) is base flow, \(\alpha\) is the filter parameter, \(Q\) is
the total streamflow, and \(k\) is the time step. A filter parameter
\(\alpha\) of 0.925 was used as in Nathan et al. (1990) and Fuka et al.
(2014). The filter was run three times (forward, backward, forward) to
attenuate the base-flow signal. Eckhardt (2023) noted that recursive
digital filters lack a physical basis, but as the method is easy to
automate, objective, and repeatable, it is appropriate for a
regional-scale study.

@@

XGBoost uses gradient boosting on decision tree algorithms to leverage
the combination of multiple very low variance models into one overall
prediction. Gradient boosting works sequentially, with the first tree
being learned on the known target values, and each subsequent tree being
learned on the errors of the tree prior. The algorithm then applies a
weighting to each tree based on this boosting to determine how much the
tree will contribute towards the final output. The final prediction is
obtained by the input data traversing all \(n\) weighted trees in the
learned ensemble. The trained XGBoost model is used to predict BFI in
ungauged catchments based on geospatial and hydroclimate predictor
variables (Table~\ref{tbl-predictors}). Certain predictors were
subdivided according to their areal coverage of each basin (e.g.~land
cover was divided into 16 subdivisions).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1892}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3784}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1081}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3243}}@{}}
\caption{Basin-characteristic variables used as predictors in XGBoost
model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation & PRISM & Basin average \\
& Mean Temperature & PRISM & Basin average \\
& Average Evapotranspiration & MODIS & Basin average \\
Geospatial & Elevation & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover & NLCD & Percent areal coverage \\
& Hydrologic Soil Group & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

@@

Statisical analyses were performed on mean annual BFI values of
instrumented streamgages. We examined the presence of temporal trends in
mean annual BFI using the Mann-Kendall nonparametric trend test
(Kendall, 1970; Mann, 1945). The Mann-Kendall test was used to determine
trends in BFI and trend significance of gauged reaches.

@@

\includegraphics{images/BFI_Instrumented_20241126.png}

@@

\caption{\label{fig-instrumented-bfi}Mean annual BFI for the period of
study (1991-2020) from instrumented stream flow data.}

@@

The long-term average BFI for the 244 gauged reaches across Arizona is
illustrated in Figure~\ref{fig-instrumented-bfi} . The highest BFI
values (\textgreater0.9) are found along the Grand Canyon in
northwestern Arizona. The highly karstic geology of this region
facilitates the rapid movement of subsurface flow to surface water and
spring outlets (Chambless et al., 2023). Relatively high BFI values
(\textgreater0.8) are found at the spring-fed headwaters of the Verde
River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
These results are consistent with interpolated BFI values reported by
Wolock (2003).

@@

@@

\includegraphics{images/BFITrends_20241126.png}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.4348}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0725}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}@{}}
\caption{Comparison of trends for BFI for all sites split by various
classifications. Only sites with a significant (\(\rho \le 0.05\)) trend
are included here as established by a Mann-Kendall test for monotonic
trends across the full period of record. n is the number of sites,
n\_pos (n\_neg) is the number of sites with positive (negative) trends,
perc\_pos (perc\_neg) is the percentage of n with a positive (negative)
trend.}\label{tbl-trends}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Classification Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_neg
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_neg
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Classification Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_neg
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_neg
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Climate - Monsoon Dominated & 94 & 10 & 23 & 0.106 & 0.245 \\
Climate - Snowmelt Dominated & 150 & 19 & 17 & 0.127 & 0.113 \\
PhysRegion - Basin\&Range & 190 & 26 & 33 & 0.137 & 0.174 \\
PhysRegion - CO Plateau & 54 & 3 & 7 & 0.056 & 0.130 \\
Climate - Warm-Wet & 31 & 3 & 6 & 0.097 & 0.194 \\
Climate - Warm-Dry & 81 & 15 & 16 & 0.185 & 0.198 \\
Climate - Cool-Wet & 83 & 6 & 11 & 0.072 & 0.133 \\
Climate - Cool-Dry & 49 & 5 & 7 & 0.102 & 0.143 \\
Slope - High & 111 & 12 & 15 & 0.108 & 0.135 \\
Slope - Low & 133 & 17 & 25 & 0.128 & 0.188 \\
\end{longtable}

@@

@@

\includegraphics{images/BFI_HUC8_20241126.png}

@@

\includegraphics{images/BFI_AllStream.pdf}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-abatzoglou2018}
Abatzoglou, J. T. et al. (2018). TerraClimate, a high-resolution global
dataset of monthly climate and climatic water balance from
1958{\textendash}2015. \emph{Scientific Data}, \emph{5}(1), 170191.
\url{https://doi.org/10.1038/sdata.2017.191}

@@

@@

@@

@@

@@

>>>>

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{losuppfig}}{\newfloat{suppfig}{h}{losuppfig}[chapter]}
\floatname{suppfig}{Figure S}
\newcommand*\quartosuppfigref[1]{Figure \hyperref[#1]{S\ref{#1}}}
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\DeclareCaptionLabelFormat{quartosuppfigreflabelformat}{#1#2}
\captionsetup[suppfig]{labelformat=quartosuppfigreflabelformat}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Supplementary Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

This study develops a technique for estimating BFI in ungauged basins in
Arizona and determines the state's long-term mean BFI. Using a machine
learning model trained with the hydrogeologic characteristics of surface
water basins, we estimate mean BFI in ungauged basins over a period of
record from 1991-2020. Regional patterns and trends in base flow and BFI
of instrumented sites are subsequently investigated. These results
provide novel analyses of low flow processes in Arizona, furthering our
understanding of climate controls on consistent flows in dryland basins.

@@

The state of Arizona, located in the southwest United States, covers a
total area of 295,253 km\(^2\). Arizona is divided into two primary
physiographic provinces: the Colorado Plateau in the northeast, and the
Basin-and-Range region in the west and south. The Central Highlands is a
transition zone consisting of scattered basins separated by the
mountainous foothills of the Mogollon Rim. The Colorado Plateau is
dominated by high-elevation desert with an average elevation of 1,936
masl (6,352 ft). Mean temperatures range from -6\(^\circ\)C
(20\(^\circ\)F) to 26\(^\circ\)C (80\(^\circ\)F) and it averages 580 mm
(23 in) of precipitation. The Basin-and-Range region has a semi-arid to
arid climate with an average elevation of 490 masl (1600 ft). Average
temperature ranges from 15\(^\circ\)C (60\(^\circ\)F) to 43\(^\circ\)C
(110\(^\circ\)F) and the region averages 200 mm (8 in) of precipitation
annually. \#\#\#CITATION\#\#\#

@@

\includegraphics{images/StudyArea_20241203.png}

@@

Arizona's hydrology varies seasonally and spatially between its
physiographic regions. In the summer, localized and intense convective
precipitation events are driven by the North American Monsoon, while in
the winter, orographic precipitation comes from Pacific frontal systems
(Eastoe et al., 2019). While monsoonal precipitation can account for up
to 50\% of annual precipitation, evaporation and dry preceding soil
properties leads to most precipitation becoming runoff (Sheppard et al.,
2002). As such, 94\% of streams in Arizona are ephemeral or intermittent
(Levick et al., 2008). Much of the hydrology of Arizona is snow-melt
derived, driven by spring melt from the high-elevation Colorado Plateau
winter snowpack. While winter precipitation provides only 30\% of annual
averages, it provides the majority of water for natural reservoirs
(Sheppard et al., 2002).

@@

Daily observed streamflow data obtained from the United States
Geological Survey (USGS) National Water Information System (NWIS) were
used in this study. Streamgages were selected depending on criteria to
ensure the applicability of each site. O'Donnell et al. (2016) found
that 8-10 years of calibration data were necessary to account for
climate variability in paired watershed studies in the region. Following
this criteria, a minimum of 10 years of record was chosen. Any years
missing more than 30 days of streamflow were removed from the possible
record period. This analysis focuses on natural, streamflow-influencing
recharge, excluding streamgages affected by regulation or diversions.
These gages were identified through annual reports on water data
published by the USGS (USGS, 2010). Finally, streamgages on the Colorado
River were excluded from the dataset on the basis that they represent
managed flows shared by Colorado River Compact states. Following these
parameters, 205 USGS streamgages with acceptable periods of record were
used in this study (Figure~\ref{fig-study-area}). Periods of record
ranged from 10 to 112 years, with a median of 28 years.

Watersheds across the United States are delineated by the USGS using a
hydrologically-defined network. This system delineates the country using
hierarchical hydrologic unit codes (HUCs), where each subsequent basin
includes the digits of the enclosing basin. Here, 8-digit HUCs (HUC 8s)
are used to divide Arizona into 84 sub-basins that are fully or
partially in the state (Figure~\ref{fig-study-area}). These HUC 8
sub-basins are analogous to medium-sized river basins and are defined by
surface water characteristics.

Annual precipitation and temperature data came from the PRISM climate
group at Oregon State University at a resolution of 4 km
(\url{https://prism.oregonstate.edu;} (Daly et al., 2008)). The PRISM
dataset provides valuable insights into regional climate in ungauged
regions and has been shown to perform well across the southwestern US
(Buban et al., 2020)\textbf{.} Instead of the water year, PRISM data
uses a calendar-year format, which was adopted for consistency in the
water balance. Although this may introduce challenges in the annual
estimates due to inter-annual snow storage, the use of long-term annual
averages is likely to reduce any potential errors (Reitz et al., 2017).

@@

Base flow was calculated using a single-parameter, recursive digital
filter technique from (Nathan et al., 1990). This base-flow separation
technique is based on a recursive digital filter used in signal analysis
that separates high-frequency signals (quickflow) from low-frequency
signals (base flow) (Lyne et al., 1979). Eckhardt (2023) noted that
recursive digital filters lack a physical basis, but as the method is
easy to automate, objective, and repeatable, it is appropriate for a
regional-scale study. The Lyne-Hollick filter has been used in (Arnold
et al., 2000; Bloomfield et al., 2009; Santhi et al., 2008; Singh et
al., 2018). It takes the form of

@@

where \(b\) is base flow, \(\alpha\) is the filter parameter, \(Q\) is
the total streamflow, and \(k\) is the time step. A filter parameter
\(\alpha\) of 0.925 was used as in Nathan et al. (1990) and Fuka et al.
(2014). The filter was run three times (forward, backward, forward) to
attenuate the base-flow signal.

@@

XGBoost uses gradient boosting on decision tree algorithms to leverage
the combination of multiple very low variance models into one overall
prediction. Gradient boosting works sequentially, with the first tree
being learned on the known target values, and each subsequent tree being
learned on the errors of the tree prior. The algorithm then applies a
weighting to each tree based on this boosting to determine how much the
tree will contribute towards the final output. The final prediction is
obtained by the input data traversing all \(n\) weighted trees in the
learned ensemble. The trained XGBoost model is used to predict BFI in
ungauged catchments based on geospatial and hydroclimate predictor
variables (Table~\ref{tbl-predictors}). Certain features were subdivided
according to their areal coverage of each basin (e.g.~land cover was
divided into 16 subdivisions).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2877}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}@{}}
\caption{Basin-characteristic variables used as initial features in
XGBoost model. Starred features are maintained in the final,
dimensionality-reduced model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation* & PRISM & Basin average \\
& Mean Temperature* & PRISM & Basin average \\
& Mean Evapotranspiration* & TerraClimate & Basin average \\
Geospatial & Elevation* & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover* & NLCD & Percent areal coverage \\
& Hydrologic Soil Group* & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

\subsubsection{Feature Selection}\label{feature-selection}

Machine learning models can face issues with overfitting when provided
with a large set of predicting features. This can cause reductions in
performance on unseen data and requires greater computational power and
memory storage (Li et al., 2017). Dimensionality reduction is a powerful
tool in addressing these issues which can be divided into two broad
categories: feature extraction and feature selection. Feature extraction
involves projecting the higher dimensional dataset to a feature space
with less dimensions, though this entails creating a new set of features
that lose the physical meaning of the original feature space. On the
other hand, feature selection maintains a subset of the original
features, all of which retain their physical meaning providing better
readability and interpretability. Here, a supervised features selection
was used to reduce the number of predictors and, as a result, increased
learning performance, decreased computational costs, and reduced
overfitting.

An initial model was trained on the full feature set of 45 predictors
(Table~\ref{tbl-predictors}). We then used a feature importance method
of feature selection to remove noise and less-important features.
Feature importance scores are an approach where importance scores are
assigned to individual features indicating how much the feature
contributed (positively or negatively) to the model (Murdoch et al.,
2019). Here, we use SHAP (SHapley Additive exPlanations) values as the
method of assigning importance values to each feature (Lundberg \& Lee,
2017). The top-10 most-important features
(\textbf{?@suppfig-shap\_values}) were subset and used in training a
subsequent model which saw performance increases and reduced computing
time. This subsequent model was used as the final model in this
analysis.

@@

Statisical analyses were performed on annual BFI and base flow values of
instrumented streamgages. We examined temporal trends using the
Mann-Kendall nonparametric trend test to determine the presence of
monotonic trends (Kendall, 1970; Mann, 1945). The Mann-Kendall test was
used to determine trends in BFI and trend significance of gauged
reaches. This test is widely used in studies of this kind (Ayers et al.,
2019; Ficklin et al., 2016; Woodhouse et al., 2022).

Trends in BFI and base flow were produced using a Mann-Kendall test for
monotonic trends. This test expects that the data is non-parametric and
that the observations are not autocorrelated over time. Observations at
4 streamgages show significant autocorrelation, though only one
streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a
significant trend in BFI. Due to this autocorrelation, the streamgage
was removed from the trend analysis as the autocorrelation may inflate
the variance of the Mann-Kendall statistic leading to an
over/underestimate of the trend (Hamed \& Rao, 1998).

@@

\includegraphics{images/BFI_Instrumented_20241203.png}

@@

\caption{\label{fig-instrumented-bfi}Mean annual BFI for the period of
record from instrumented stream flow data.}

@@

The long-term average BFI for the 205 gauged reaches across Arizona is
illustrated in Figure~\ref{fig-instrumented-bfi} . The highest BFI
values (\textgreater0.9) are found along the Grand Canyon in
northwestern Arizona. The highly karstic geology of this region
facilitates the rapid movement of subsurface flow to surface water and
spring outlets (Chambless et al., 2023). Relatively high BFI values
(\textgreater0.8) are found at the spring-fed headwaters of the Verde
River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
These results are consistent with interpolated BFI values reported by
Wolock (2003).

@@

Trends in BFI over the period of record for each streamgage in this
analysis are illustrated in Figure~\ref{fig-instrumented-trend} and
Table~\ref{tbl-trends} .

@@

\includegraphics{images/BFITrends_20241203.png}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.4800}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0933}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0933}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1333}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1333}}@{}}
\caption{Comparison of trends for BFI for all sites split by various
classifications. Only sites with a significant (\(\rho \le 0.05\)) trend
are included here as established by a Mann-Kendall test for monotonic
trends across the full period of record. n is the number of sites,
n\_pos (n\_neg) is the number of sites with positive (negative) trends,
perc\_pos (perc\_neg) is the percentage of n with a positive (negative)
trend.}\label{tbl-trends}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Classification Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n\_neg
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
perc\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
perc\_neg
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Classification Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n\_neg
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
perc\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
perc\_neg
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Precipitation - Monsoon Dominated & 87 & 8 & 21 & 0.092 & 0.241 \\
Precipitation - Snowmelt Dominated & 118 & 9 & 12 & 0.076 & 0.102 \\
PhysRegion - Basin\&Range & 156 & 14 & 26 & 0.090 & 0.167 \\
PhysRegion - CO Plateau & 49 & 3 & 7 & 0.061 & 0.143 \\
Climate - Warm-Wet & 31 & 2 & 6 & 0.065 & 0.194 \\
Climate - Warm-Dry & 55 & 6 & 11 & 0.109 & 0.200 \\
Climate - Cool-Wet & 74 & 4 & 9 & 0.054 & 0.122 \\
Climate - Cool-Dry & 45 & 5 & 7 & 0.111 & 0.156 \\
Slope - High & 102 & 10 & 12 & 0.098 & 0.118 \\
Slope - Low & 103 & 7 & 21 & 0.068 & 0.204 \\
\end{longtable}

@@

\subsubsection{Predictor Importance}\label{predictor-importance}

\includegraphics{images/actual-predicted.png}

\caption{\label{fig-actual_predicted}Linear relationship between
observed BFI and predicted BFI. The solid line is the 1:1 line, the
dashed line is regressed to the data.}

@@

\includegraphics{images/BFI_HUC8_20241203.png}

@@

\includegraphics{images/BFI_AllStream.png}

@@

\section{Supporting Information}\label{supporting-information}

\includegraphics{images/supplemental/shap_summary.png}
\{\#suppfig-shap\_values\}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abatzoglou2018}
Abatzoglou, J. T. et al. (2018). TerraClimate, a high-resolution global
dataset of monthly climate and climatic water balance from
1958{\textendash}2015. \emph{Scientific Data}, \emph{5}(1), 170191.
\url{https://doi.org/10.1038/sdata.2017.191}

@@

\bibitem[\citeproctext]{ref-ayers2019}
Ayers, J. R., Villarini, G., Jones, C., \& Schilling, K. (2019). Changes
in monthly baseflow across the U.S. Midwest. \emph{Hydrological
Processes}, \emph{33}(5), 748--758.
\url{https://doi.org/10.1002/hyp.13359}

@@

\bibitem[\citeproctext]{ref-MK_Hamed1998}
Hamed, K. H., \& Rao, A. (1998). A modified mann-kendall trend test for
autocorrelated data. \emph{Journal of Hydrology}, \emph{204}(1),
182--196.
https://doi.org/\url{https://doi.org/10.1016/S0022-1694(97)00125-X}

@@

\bibitem[\citeproctext]{ref-feat_selec2017}
Li, J., Cheng, K., Wang, S., Morstatter, F., Trevino, R. P., Tang, J.,
\& Liu, H. (2017). Feature selection: A data perspective. \emph{ACM
Comput. Surv.}, \emph{50}(6). \url{https://doi.org/10.1145/3136625}

\bibitem[\citeproctext]{ref-SHAP_values}
Lundberg, S. M., \& Lee, S.-I. (2017). A unified approach to
interpreting model predictions, \emph{30}. Retrieved from
\url{https://proceedings.neurips.cc/paper_files/paper/2017/file/8a20a8621978632d76c43dfd28b67767-Paper.pdf}

@@

\bibitem[\citeproctext]{ref-ML_interpretability2019}
Murdoch, W. J., Singh, C., Kumbier, K., Abbasi-Asl, R., \& Yu, B.
(2019). Definitions, methods, and applications in interpretable machine
learning. \emph{Proceedings of the National Academy of Sciences},
\emph{116}(44), 22071--22080.
\url{https://doi.org/10.1073/pnas.1900654116}

@@

\bibitem[\citeproctext]{ref-woodhouse_udall22}
Woodhouse, C. A. et al. (2022). Upper {Gila}, {Salt}, and {Verde
Rivers}: {Arid Land Rivers} in a {Changing Climate}. \emph{Earth
Interactions}, \emph{26}(1), 1--14.
\url{https://doi.org/10.1175/EI-D-21-0014.1}
